Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2015-01-25 algorithm.tex
AA: half reading

At a glance, at each iteration simulations are performed. Regions are classified based on the innovation introduced by their configurations in the Pareto-front so far calculated. The most interesting regions are split, so that each sub-region will receive evaluation effort in the next iteration. On the contrary, uninteresting regions are merged.

@@

For $era_{i}$ with $i>0$ the following operations must be performed.
\begin{enumerate}
\item \label{pers02.enu:K}
In each era $i$, a set $test_i$ of $K$ configurations will be evaluated. We call them \emph{test configurations} and selecting them is exactly the goal of this step.
We select only \emph{new} configurations, i.e. the ones that have not been evaluated in past eras. In other words, $test_i$ must not contain $\bigcup_{j=0}^{i-1}test_{j}$.
Since all regions $R$ in the configuration space partition $\mathcal{R}_{i}$ must be properly explored, we have to distribute the test configurations among all these regions. For this reason, $\frac{K}{\left|\mathcal{R}_{i}\right|}$ configurations are randomly extracted from the new configurations of each region $R\in\mathcal{R}_{i}$. We use $test_{R,i}$ to denote the set of the selected configurations and, obviously, $\bigcup_{R\in\mathcal{R}_{i}} test_{R,i} = test_i$.
 In case the new configurations in $R$ are less then $\frac{K}{\left|\mathcal{R}_{i}\right|}$, all of them are selected. In this case $test_{R,i}=R\setminus\bigcup_{j=0}^{i-1}test_{j}$, where  $\bigcup_{j=0}^{i-1}test_{j}$ is the set of the configurations already in the past.\footnote{Note that if such regions exist, the total number of evaluations in era $i$ will be less than $K$}
\item All configurations on $test_{i}$ are simulated.
\item The Pareto set $\mathscr{P}_{i}$ for $era_{i}$ is defined as the set of non-dominated configurations among the ones evaluated so far, i.e.
	\[
	\mathscr{P}_{i} \triangleq
	\mathscr{P}\left(\bigcup_{j\leq i}test_j\right)=
	\mathscr{P}\left(test_{i}\cup\mathscr{P}_{i-1}\right)
	\]
where the last equality can easily be proved.

@@

\item \label{enu:region_innovation_score} Every region $R\in\mathcal{R}_{i}$ is given an innovation score defined
as the sum of the innovation scores of its non-dominated test configurations:
	\[
	is\left(R\right)=\sum\left\{ \left.is\left(\mathbf{c}\right)\right|\mathbf{c}\in\mathscr{P}_{i}\cap R\right\} 
	\]

@@

\item The set of regions $\mathcal{R}_{i+1}$ for the following era is composed of the sub-regions resulting from splitting operations ($\psi\left(\mathscr{R}_{i,h}\right)$), the low innovation regions unchanged $\mathcal{R}_{i,l}$, the result of the merging operations ($\mu(\mathscr{M}_i))$ and the no innovation regions that was not possible to merge $\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)$. Formally:
	\begin{align} 
	\mathcal{R}_{i+1} \triangleq
	\psi\left(\mathscr{R}_{i,h}\right) \cup \mathcal{R}_{i,l}\cup
	\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)\cup \mu(\mathscr{M}_i)
	\end{align}
\end{enumerate}

@@

			\nl simulate ($test_i$);\\
			\nl $\mathscr{P}_i \Leftarrow 
				\mathscr{P}\left(test_{i}\cup\mathscr{P}_{i-1}\right)$;\\
			\nl \ForEach{$\mathbf{c}^*\in\mathscr{P}_i$} {
				\nl $is\left(\mathbf{c}^*\right) \Leftarrow \min\left\{ \left.s\left(\mathbf{c}\rightarrow\mathbf{c} \right)\right|\mathbf{c}\in\mathscr{P}_{i-1}\right\}$;\\
				}
			\nl \ForEach{$R\in\mathscr{R}_i$} {
				\nl $is\left(R\right)=\sum\left\{ \left.is\left(\mathbf{c}\right)\right|\mathbf{c}\in\mathscr{P}_{i}\cap R\right\} $;
			}

@@

\subsection{Termination Condition}
The algorithm terminates at step $k$ if
\[
is_{TOT,k},is_{TOT,k-1},\dots,is_{TOT,k-\beta}\le\gamma\cdot is_{TOT,k-\left(\beta-1\right)}
\]
 where $\beta$ and $\gamma$ are selected by the experimenter. 
The algorithm terminates when in the last
$\beta$ iterations not a great ``amount of innovation'' has been
discovered. Therefore it is very probably that carrying on iterating
will not considerably change the Pareto-front approximation already formed.

It is not recommended to chose $\beta=1$, because it is possible that the
$k-1$-th and $k$-th Pareto front approximations are not very different but next ones could be. In other words,
it is unsafe to terminate as soon as the Pareto front does not considerably change
between two consecutive eras, thus a strategy taking into
account the history of the Pareto fronts has been adopted in the
proposed approach.

@@

\subsection{Exploration vs Exploitation in PS}
The proposed algorithm has been designed in order to be inherently able to balance exploration and exploitation, which are ``two cornerstones of problem solving by search''~\cite{eiben1998evolutionary}. While a detailed survey on the exploration vs exploitation trade-off can be found in~\cite{vcrepinvsek2013exploration}, here we limitedly recall that \emph{exploration} is the capability of an algorithm to consider different parts of the configuration space, not being confined only to a limited subset and being trapped in local optima. Following the terminology of \cite{weise2012evolutionary}, if the exploration is not sufficiently wide, the algorithm suffers from \emph{premature convergence}, i.e. it starts to analyze only a narrow portion of the configuration space, thus ignoring potential Pareto configurations that are outside this portion. When premature convergence happens, there is insufficient \emph{diversity}~\cite{weise2012evolutionary}, i.e. all the provided solutions are similar.
On the other hand, \emph{exploitation} is the capability to recognize and densely analyze the parts of configuration space that have many Pareto configurations.

>>>>

At a glance, at each iteration simulations are performed. Regions are classified based on the innovation introduced by their configurations in the Pareto-front so far calculated. The most interesting regions are split, so that each sub-region will receive more evaluation effort in the next iteration. On the contrary, uninteresting regions are merged.

@@

For $era_{i}$ with $i>0$ the following operations must be performed.
\begin{enumerate}
\item \label{pers02.enu:K}
In each era $i$, a set $test_i$ of $K$ configurations will be evaluated. We call them \emph{test configurations} and selecting them is exactly the goal of this step.
We select only \emph{new} configurations, i.e. the ones that have not been evaluated in past eras. In other words, $test_i$ must not overlap with $\bigcup_{j=0}^{i-1}test_{j}$.
Since all regions $R$ in the configuration space partition $\mathcal{R}_{i}$ must be properly explored, we have to distribute the test configurations among all these regions. For this reason, $\frac{K}{\left|\mathcal{R}_{i}\right|}$ configurations are randomly extracted from the new configurations of each region $R\in\mathcal{R}_{i}$. We use $test_{R,i}$ to denote the set of the selected configurations and, obviously, $\bigcup_{R\in\mathcal{R}_{i}} test_{R,i} = test_i$.
 In case the new configurations in $R$ are less then $\frac{K}{\left|\mathcal{R}_{i}\right|}$, all of them are selected. In this case $test_{R,i}=R\setminus\bigcup_{j=0}^{i-1}test_{j}$, where  $\bigcup_{j=0}^{i-1}test_{j}$ is the set of the configurations already evaluated in the past.\footnote{Note that if such regions exist, the total number of evaluations in era $i$ will be less than $K$}
\item All configurations on $test_{i}$ are simulated.
\item The Pareto set $\mathscr{P}_{i}$ for $era_{i}$ is defined as the set of non-dominated configurations among the ones evaluated so far, i.e.
	\[
	\mathscr{P}_{i} \triangleq
	\mathscr{P}\left(\bigcup_{j\leq i}test_j\right)=
	\mathscr{P}\left(test_{i}\cup\mathscr{P}_{i-1}\right)
	\]
where the last equality can easily be proved.

@@

\item \label{enu:region_innovation_score} Every region $R\in\mathcal{R}_{i}$ is given an innovation score defined
as the sum of the innovation scores of its non-dominated test configurations:
	\[
	is\left(R\right)=\sum\left\{ \left.is\left(\mathbf{c}\right)\right|\mathbf{c}\in\mathscr{P}_{i}\cap test_{R,i} \right\} 
	\]

@@

\item The set of regions $\mathcal{R}_{i+1}$ for the following era is composed of the sub-regions resulting from splitting operations ($\psi\left(\mathscr{R}_{i,h}\right)$), the low innovation regions unchanged $\mathcal{R}_{i,l}$, the result of the merging operations ($\mu(\mathscr{M}_i))$ and the no innovation regions that was not possible to merge $\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathscr{M}_i }\left\{ R_{1},R_{2}\right\} \right)$. Formally:
	\begin{align} 
	\mathcal{R}_{i+1} \triangleq
	\psi\left(\mathscr{R}_{i,h}\right) \cup \mathcal{R}_{i,l}\cup
	\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathscr{M}_i }\left\{ R_{1},R_{2}\right\} \right)\cup \mu(\mathscr{M}_i)
	\end{align}
\end{enumerate}

@@

			\nl simulate ($test_i$);\\
			\nl $\mathscr{P}_i \Leftarrow 
				\mathscr{P}\left(test_{i}\cup\mathscr{P}_{i-1}\right)$;\\
			\nl \ForEach{$\mathbf{c}^*\in\mathscr{P}_i$} {
				\nl $is\left(\mathbf{c}^*\right) \Leftarrow \min\left\{ \left.s\left(\mathbf{c}\rightarrow\mathbf{c} \right)\right|\mathbf{c}\in\mathscr{P}_{i-1}\right\}$;\\
				}
			\nl \ForEach{$R\in\mathscr{R}_i$} {
				\nl $is\left(R\right)=\sum\left\{ \left.is\left(\mathbf{c}\right)\right|\mathbf{c}\in\mathscr{P}_{i}\cap test_{R,i} \right\} $;
			}

@@

\subsection{Termination Condition}
The algorithm terminates at step $k$ if
\[
is_{TOT,k},is_{TOT,k-1},\dots,is_{TOT,k-\beta}\le\gamma\cdot is_{TOT,k-\left(\beta-1\right)}
\]
 where $\beta$ and $\gamma$ are selected by the experimenter. 
In other words, the algorithm terminates if, in the last
$\beta$ iterations, the ``amount of innovation'' has been
small. Therefore, it is very probably that carrying on iterating
will not considerably change the Pareto-front approximation already formed.

It is not recommended to chose $\beta=1$, because it is possible that the
$k-1$-th and $k$-th Pareto front approximations are not very different but next ones could be. In other words,
it is unsafe to terminate as soon as the Pareto front does not considerably change
between two consecutive eras. For this reason, we take into
account the history of the Pareto fronts.

@@

\subsection{Exploration vs. Exploitation in PS}
The proposed algorithm has been designed in order to be inherently able to balance exploration and exploitation, which are ``two cornerstones of problem solving by search''~\cite{eiben1998evolutionary}. While a detailed survey on the exploration vs. exploitation trade-off can be found in~\cite{vcrepinvsek2013exploration}, here we limitedly recall that \emph{exploration} is the capability of an algorithm to consider different parts of the configuration space, not being confined only to a limited subset and being trapped in local optima. Following the terminology of \cite{weise2012evolutionary}, if the exploration is not sufficiently wide, the algorithm suffers from \emph{premature convergence}, i.e. it starts to analyze only a narrow portion of the configuration space, thus ignoring potential Pareto configurations that are outside this portion. When premature convergence happens, there is insufficient \emph{diversity}~\cite{weise2012evolutionary}, i.e. all the provided solutions are similar.
On the other hand, \emph{exploitation} is the capability to recognize and densely analyze the parts of configuration space that have many Pareto configurations.

### 2015-01-25 experiments.tex
AA: half reading

\figR{dispersion}(c) shows the average normalized absolute dispersion
errors for different benchmarks for PS and GA exploration.  As it can
be observed, GA shows a tendency in discovering more uniformly
dispersed points (i.e. higher values in the chart). These results,
together with the variation ranges of \figR{dispersion}(a) and (b)
confirm the of density vs extension trade-offs when comparing PS and
GA approaches. We highlight that PS shows a stronger \emph{uniform convergence} (discussed in \secR{metrics}), i.e. it provides a more extended Pareto front, avoiding the designer to neglect some important options, and offering a wider view of the potentiality of the designed device.

>>>>

\figR{dispersion}(c) shows the average normalized absolute dispersion
errors for different benchmarks for PS and GA exploration.  As it can
be observed, GA shows a tendency in discovering more uniformly
dispersed points (i.e. higher values in the chart). These results,
together with the variation ranges of \figR{dispersion}(a) and (b)
confirm the of density vs. extension trade-offs when comparing PS and
GA approaches. We highlight that PS shows a stronger \emph{uniform convergence} (discussed in \secR{metrics}), i.e. it provides a more extended Pareto front, avoiding the designer to neglect some important options, and offering a wider view of the potentiality of the designed device.
